---
title: "Assumptions in Causal Inference"
author: "Lucy D'Agostino McGowan"
format: kakashi-revealjs
footer: "Slides by [Dr. Lucy D'Agostino McGowan](https://lucymcgowan.com)"
---


## {background-color="#23373B" .center}

1. **Consistency** 
2. **Exchangeability**
3. **Positivity**

<!-- ## {{< fa laptop >}} Application Exercise -->

<!-- ::: nonincremental -->
<!-- ::: small -->
<!-- 1. Create a new project using my template here: -->

<!-- ```bash -->
<!-- https://github.com/sta-779-f23/appex-02.git -->
<!-- ``` -->

<!-- 2. Using the data below with two potential outcomes, `y_0` and `y_1` calculate the true causal effect for each individual as well as the true average causal effect for these ten individuals. -->

<!-- ```{r} -->
<!-- data <- data.frame( -->
<!--   id = 1:10, -->
<!--   y_1 = c(4, 4, 6, 5, 6, 5, 6, 7, 5, 6), -->
<!--   y_0 = c(1, 3, 4, 5, 5, 6, 8, 6, 3, 5) -->
<!-- ) -->
<!-- ``` -->

<!-- ::: -->
<!-- ::: -->

## Consistency

* We assume that the causal question you claim you are answering is consistent with the one you are *actually* answering with your analysis.
* Mathematically: $Y_{obs} = (X)Y(1) + (1-X)Y(0)$
* **Well defined exposure**
* **No interference** 

## Well defined exposure

* We assume that for each value of the exposure, there is no difference between subjects in the delivery of that exposure
* Put another way, multiple versions of the treatment do not exist

## {{< fa pencil >}} Application exercise

:::nonincremental
1. Think of an example where this might be violated
2. Turn to the person next to you and tell them your example
3. Pick your favorite between the two of you, we'll share them with the class
:::

```{r}
#| echo: false
countdown::countdown(3)
```

## No interference

* We assume that the outcome (technically all *potential* outcomes, regardless of whether they are observed) for any subject does not depend on another subject's exposure

## {{< fa pencil >}} Application exercise

:::nonincremental
1. Think of an example where this might be violated
2. Turn to the person next to you and tell them your example
3. Pick your favorite between the two of you, we'll share them with the class
:::

```{r}
#| echo: false
countdown::countdown(3)
```


## Exchangeability

* We assume that within levels of relevant variables (confounders), exposed and unexposed subjects have an equal likelihood of experiencing any outcome prior to exposure
* i.e. the exposed and unexposed subjects are exchangeable
* This assumption is sometimes referred to as **no unmeasured confounding**.

## {{< fa pencil >}} Application exercise

:::nonincremental
1. Think of an example where this might be violated
2. Turn to the person next to you and tell them your example
3. Pick your favorite between the two of you, we'll share them with the class
:::

```{r}
#| echo: false
countdown::countdown(3)
```


## Positivity

* We assume that within each level and combination of the study variables used to achieve exchangeability, there are exposed and unexposed subjects. 
* Said differently, each individual has some chance of experiencing every available exposure level.
* Sometimes this is referred to as the **probabilistic** assumption.

## {{< fa pencil >}} Application exercise

:::nonincremental
1. Think of an example where this might be violated
2. Turn to the person next to you and tell them your example
3. Pick your favorite between the two of you, we'll share them with the class
:::

```{r}
#| echo: false
countdown::countdown(3)
```



