---
title: "From question to answer: stratification and outcome models"
author: "Lucy D'Agostino McGowan"
format: revealjs
---


#  Causal inference with `group_by()` and `summarize()` {background-color="#533146" .center}

```{r}
#| label: setup
#| include: false
source(here::here("setup.R"))
library(tidyverse)
library(ggdag)

set.seed(1234)
```


## Example

::: small
* We are analyzing data from a software company 
* Effect: Software update frequency (`weekly` or `daily`) on customer satisfaction
* Free customers: more likely to receive weekly updates
* Premium customers: more likely to receive daily updates
* Premium customers: more likely to have higher satisfaction
:::

## Example DAG

::: {.box .fragment}
Are there any open backdoor paths?
:::



```{r}
#| code-fold: true
#| fig-align: center

coords1 <- list(
  x = c(customer_type = 1, updates = 2, satisfaction = 3),
  y = c(customer_type = 0, updates = 0, satisfaction = 0)
)

dag1 <- dagify(
  satisfaction ~ customer_type,
  updates ~ customer_type,
  coords = coords1,
  labels = c(
    customer_type = "customer type",
    updates = "frequency\nof updates",
    satisfaction = "customer\nsatisfaction"
  )
)

ggdag(dag1, use_text = FALSE, use_edges = FALSE) +
  geom_dag_text(aes(label = label), nudge_y = c(-.05, -.05, -.05), color = "black") +
  geom_dag_edges_arc(curvature = c(0.07, 0)) +
  theme_dag() +
  ylim(c(.2, -.2))
```

## Example Data

```{r}
#| code-fold: true
set.seed(1)
n <- 10000
satisfaction1 <- tibble(
  # Free (0) or Premium (1)
  customer_type = rbinom(n, 1, 0.5),
  p_exposure = case_when(
    # Premium customers are more likely to receive daily updates
    customer_type == 1 ~ 0.75,
    # Free customers are more likely to receive weekly updates
    customer_type == 0 ~ 0.25
  ),
  # Weekly (0) vs Daily (1)
  update_frequency = rbinom(n, 1, p_exposure),
  # generate the "true" average treatment effect of 0
  # to do this, we are going to generate the
  # potential outcomes, first if exposure = 0
  # `y0` = `satisfaction(weekly)`
  # notice `update_frequency` is not in the equation below
  # we use rnorm(n) to add the random error term that is normally
  # distributed with a mean of 0 and a standard deviation of 1
  y0 = customer_type + rnorm(n),
  # because the true effect is 0, the potential outcome
  # if exposure = 1 is identical
  y1 = y0,
  # in practice, we will only see one of these
  # observed
  satisfaction = (1 - update_frequency) * y0 +
    update_frequency * y1,
  observed_potential_outcome = case_when(
    update_frequency == 0 ~ "y0",
    update_frequency == 1 ~ "y1"
  )
) |>
  mutate(
    update_frequency = factor(
      update_frequency,
      labels = c("weekly", "daily")
    ),
    customer_type = factor(
      customer_type,
      labels = c("free", "premium")
    )
  )
satisfaction1 |>
  select(update_frequency, customer_type, satisfaction)
```

## Example analysis

::: {.box .fragment}
What assumptions am I making with this analysis?
:::

```{r}
satisfaction1 |>
  group_by(update_frequency) |>
  summarise(avg_satisfaction = mean(satisfaction))
```

::: {.columns layout-ncol=3}

::: {column}
* Exchangeability

:::

::: {column}
* Consistency

:::

::: {column}
* Positivity

:::

:::

## Example analysis (stratified)

::: box
We know the exchangeability assumption is violated (I simulated it as such!)

:::

::: {.fragment}

```{r}
satisfaction_strat <- satisfaction1 |>
  group_by(customer_type, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )

satisfaction_strat
```

:::

## Example analysis (stratified)

```{r}
satisfaction_strat_est <- satisfaction_strat |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = daily - weekly)

satisfaction_strat_est
```


## Example analysis (stratified)

```{r}
satisfaction_strat_est |>
  # note: we would need to weight this if the confounder
  # groups were not equally sized
  summarise(estimate = mean(estimate))
```

::: {.box .fragment}

What is my estimated causal effect?

:::

## Example 2 

* Now we have **two binary confounders**
* Weekly updates have a higher chance of occurring within business hours
* Daily updates have a higher chance of occurring after business hours
* Customers who don’t overlap with business hours have lower satisfaction 

## Example 2 DAG

::: {.box .fragment}
Are there any open backdoor paths?
:::

```{r}
#| code-fold: true
#| fig-align: center
dag2 <- dagify(
  satisfaction ~ customer_service + customer_type,
  customer_service ~ business_hours,
  updates ~ customer_type + business_hours,
  coords = time_ordered_coords(),
  labels = c(
    customer_type = "customer\ntype",
    business_hours = "business\nhours",
    updates = "frequency\nof updates",
    customer_service = "customer\nservice",
    satisfaction = "customer\nsatisfaction"
  )
)

ggdag(dag2, use_text = FALSE) +
  geom_dag_text(
    aes(label = label),
    nudge_y = c(-.35, -.35, .35, .35, .35),
    color = "black"
  ) +
  theme_dag()
```

## Example 2 Data

```{r}
#| code-fold: true
satisfaction2 <- tibble(
  # Free (0) or Premium (1)
  customer_type = rbinom(n, 1, 0.5),
  # Business hours (Yes: 1, No: 0)
  business_hours = rbinom(n, 1, 0.5),
  p_exposure = case_when(
    customer_type == 1 & business_hours == 1 ~ 0.75,
    customer_type == 0 & business_hours == 1 ~ 0.9,
    customer_type == 1 & business_hours == 0 ~ 0.2,
    customer_type == 0 & business_hours == 0 ~ 0.1
  ),
  # Weekly (0) vs Daily (1)
  update_frequency = rbinom(n, 1, p_exposure),
  # More likely during business hours
  customer_service_prob = business_hours * 0.9 +
    (1 - business_hours) * 0.2,
  customer_service = rbinom(n, 1, prob = customer_service_prob),
  satisfaction = 70 + 10 * customer_type +
    15 * customer_service + rnorm(n),
) |>
  mutate(
    customer_type = factor(
      customer_type,
      labels = c("free", "premium")
    ),
    business_hours = factor(
      business_hours,
      labels = c("no", "yes")
    ),
    update_frequency = factor(
      update_frequency,
      labels = c("weekly", "daily")
    ),
    customer_service = factor(
      customer_service,
      labels = c("no", "yes")
    )
  )

satisfaction2 |>
  select(update_frequency, customer_type, business_hours, customer_service, satisfaction) |>
  print(n = 5)
```

## Example 2 Analysis

```{r}
satisfaction2_strat <- satisfaction2 |>
  group_by(customer_type, business_hours, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )

satisfaction2_strat |>
  select(avg_satisfaction, everything())
```

::: {.box .fragment}

What did I "adjust" for?

:::

## Example 2 Analysis

```{r}
satisfaction2_strat |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

::: {.box .fragment}
What is my estimated causal effect?
:::

## Example 2 Analysis

```{r}
satisfaction2_strat <- satisfaction2 |>
  group_by(customer_type, customer_service, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )

satisfaction2_strat |>
  select(avg_satisfaction, everything())
```

::: {.box .fragment}

What did I "adjust" for?

:::

## Example 2 Analysis

```{r}
satisfaction2_strat |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

::: {.box .fragment}
What is my estimated causal effect?
:::

## Example 3

* What if I have a **continuous confounder**
* The number of users within the organization
* Organizations with more users get more updates and have slightly lower satisfaction scores.

## Example 3 DAG


```{r}
#| code-fold: true
#| fig-align: center

coords3 <- list(
  x = c(num_users = 1, updates = 2, satisfaction = 3),
  y = c(num_users = 0, updates = 0, satisfaction = 0)
)

dag3 <- dagify(
  satisfaction ~ num_users,
  updates ~ num_users,
  coords = coords3,
  labels = c(
    num_users = "number of\nusers",
    updates = "frequency\nof updates",
    satisfaction = "customer\nsatisfaction"
  )
)

ggdag(dag3, use_text = FALSE, use_edges = FALSE) +
  geom_dag_text(aes(label = label), nudge_y = c(-.05, -.05, -.05), color = "black") +
  geom_dag_edges_arc(curvature = c(0.07, 0)) +
  theme_dag() +
  ylim(c(.2, -.2))
```


## Example 3 data

```{r}
#| code-fold: true

satisfaction3 <- tibble(
  # Number of users
  num_users = runif(n, 1, 500),
  # Larger customers more likely to have daily updates
  update_frequency = rbinom(n, 1, plogis(num_users / 100)),
  # with more users come less satisfaction
  satisfaction = 70 + -0.2 * num_users
) |>
  mutate(
    update_frequency = factor(
      update_frequency,
      labels = c("weekly", "daily")
    )
  )

satisfaction3
```

## Example 3 analysis

```{r}
satisfaction3_strat <- satisfaction3 |>
  mutate(num_users_q = ntile(num_users, 5)) |>
  group_by(num_users_q, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )

satisfaction3_strat
```

## Example 3 analysis


```{r}
satisfaction3_strat |>
  ungroup() |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

::: {.box .fragment}
What is my estimated causal effect?
:::


## Example 3 analysis

```{r}
#| code-line-numbers: "|2"
satisfaction3_strat <- satisfaction3 |>
  mutate(num_users_q = ntile(num_users, 10)) |>
  group_by(num_users_q, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )
```

## Example 3 analysis


```{r}
satisfaction3_strat |>
  ungroup() |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

## Example 3 analysis

```{r}
#| code-line-numbers: "2"
satisfaction3_strat <- satisfaction3 |>
  mutate(num_users_q = ntile(num_users, 30)) |>
  group_by(num_users_q, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )
```

## Example 3 analysis


```{r}
satisfaction3_strat |>
  ungroup() |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

::: {.box .fragment}
What is my estimated causal effect?
:::

## Example 3 analysis

```{r}
#| code-line-numbers: "2"
satisfaction3_strat <- satisfaction3 |>
  mutate(num_users_q = ntile(num_users, 100)) |>
  group_by(num_users_q, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )
```

## Example 3 analysis


```{r}
satisfaction3_strat |>
  ungroup() |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```

::: {.box .fragment}
What is my estimated causal effect?
:::

## Stratification

* The approach we’ve been using with `group_by()` and `summarize()` is often called **stratification** 
* This is a non-parametric (ish) approach
* Can be powerful for simple problems or when you have lots of data because you can sometimes avoid model misspecification problems
* With many confounders (especially continuous ones), we quickly encounter the curse of dimensionality
  * (i.e. too few observations by combinations of confounder levels)

# Parametric outcome models {background-color="#533146" .center}

## Conditional means

* You can think of **stratification** as calculating conditional means
  * We then marginalized over these by taking a (weighted) average of these averages to get the overall estimated average treatment effect 
* A more general extension of conditional means is **multivariable linear regression**

## Multivariable linear regression

* Also called: 
    * Direct adjustment
    * Regression adjustment
    
## Example 2 analysis (regression)

```{r}
#| code-line-numbers: "|2,3,4,5"
library(broom)
lm(
  satisfaction ~ update_frequency + customer_type + business_hours,
  data = satisfaction2
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

## Example 3 analysis (regression)

```{r}
lm(
  satisfaction ~ update_frequency + num_users,
  data = satisfaction3
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

## Parametric Models

* This generalization doesn’t come for free!
* We’ve now introduced a **parametric statistical model** to make estimates across the regions of our data
* Our estimate in the continuous case gives us exactly the right answer because the statistical model under `lm()` exactly matches our simulation

## Example 4

```{r}
#| code-fold: true
#| fig-align: center

set.seed(1)
satisfaction4 <- tibble(
  # Number of users
  num_users = runif(n, 1, 500),
  # Larger customers more likely to have daily updates
  update_frequency = rbinom(n, 1, plogis(num_users / 100)),
  # with more users come less satisfaction
  satisfaction = 70 - 0.0001 * (num_users-250)^2 - 0.0001 * (num_users - 250)^3
) |>
  mutate(
    update_frequency = factor(
      update_frequency,
      labels = c("weekly", "daily")
    )
  )

ggplot(satisfaction4, aes(x = num_users, y = satisfaction)) +
  geom_point()
```
## Example 4 analysis (regression)

```{r}
lm(
  satisfaction ~ update_frequency + num_users,
  data = satisfaction4
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

## Example 4 analysis (stratified)

```{r}
satisfaction4_strat <- satisfaction4 |>
  mutate(num_users_q = ntile(num_users, 40)) |>
  group_by(num_users_q, update_frequency) |>
  summarise(
    avg_satisfaction = mean(satisfaction),
    .groups = "drop"
  )

satisfaction4_strat |>
  ungroup() |>
  pivot_wider(
    names_from = update_frequency,
    values_from = avg_satisfaction
  ) |>
  summarise(estimate = mean(daily - weekly))
```
## Example 4 analysis (regression, polynomial)

```{r}
#| code-line-numbers: "|2"
lm(
  satisfaction ~ update_frequency + poly(num_users, 3),
  data = satisfaction4
) |>
  tidy(conf.int = TRUE) |>
  filter(term == "update_frequencydaily") |>
  select(estimate, starts_with("conf"))
```

## Regression

* Outcome regression can work very well when we meet the assumptions of the estimator we’re using for our model
* OLS, for instance, can be very beneficial if we understand the relationships between the outcome and the variables in the regression

## Regression

* In this example, OLS did work when we had a linear confounder
* It did not work when we had a nonlinear confounder but modelled it linearly
* It did work when we had a nonlinear confounder and modelled it correctly

## Conditional vs marginal effects

* Outcome models give us **conditional effects**
* "a one-unit change in the exposure results in a `coefficient` change in the outcome *holding all other variables in the model constant*" 
* In causal inference, we are often interested in **marginal effects**

## Marginal effects

* We want to average the effect across the distribution of factors in a particular population for which we are trying to estimate a causal effect 

## Marginal effects

**The estimates will be identical when:**

  * the outcome is continuous AND
  * the effect is linear AND
  * there are no interactions between the exposure effect and other factors about the population
  
## Marginal effects

* What if ther is an interaction in the model?
  * that is, what if the exposure has a different impact on the outcome depending on some other factor
* Now we no longer have a single coefficient to interpret 
* We may want to estimate a marginal effect, taking into account the distribution of that factor in the population of interest

## Marginal effects

* Why? We are ultimately trying to determine whether we should suggest exposure to the target population, so we want to know, on average, whether it will be beneficial

## Example 5

* Consider a variation of our first example where update frequency **does** have a causal effect, but that effect varies by customer type
* For premium customers, daily updates **increase** satisfaction by 5 points
* For free customers, daily updates **decrease** satisfaction by 5 points

## Example 5

:::: {.columns}

::: column

### 50% of the customers are premium and 50% are free

::: {.fragment}

`(.5 * 5) + (.5 * -5)=`

::: large
**0**
:::

:::

:::


::: column

###  100% of the customers are premium

::: {.fragment}

`(1 * 5) + (0 * -5) =`

::: large
**5**
:::

:::

:::

::: column

### 100% of the customers are free

::: {.fragment}

`(0 * 5) + (1 * -5) =`

::: large
**-5**
:::

:::

:::

::::



